%{{{ latex prelude
\documentclass{article}
\usepackage{xcolor}

\def\fb#1{\textbf{#1}}

\newcommand{\todo}[1]{[\textcolor{red}{TODO}: #1]}

\title{Roles in Agent Oriented Programming\\\normalsize{a survey}}
\author{Claudia Grigore, Rem Collier, Radu Grigore}
% latex prelude }}}

\begin{document}
\maketitle

% {{{ abstract
\begin{abstract}
\end{abstract}
% abstract }}}
%{{{ introduction
\section{Introduction}

AOP (\fb agent \fb oriented \fb programming) is a research area
whose goal is to find better ways to design, analyse, and develop
programs by using concepts imported from sociology. The products
of researchers working in this area range from methodologies that
are useful in designing programs to programming languages. The
type of applications that benefit the most from these results
are those that are inherently distributed, an example being,
unsurprisingly, simulators of societies.

It is fairly clear now how certain social concepts,
such as belief and behaviour, apply to programming
(Section~\ref{sec:background}). Much recent work focuses on
the assimilation from sociology of the concepts of role and
organisation, which are closely related.

\todo{Continue here.}
Clearly, AOP is inspired by sociology. Recently there has been
much interest in importing other concepts from sociology, such
as roles and organisations, in order to make it easier to write
large scale programs. In this paper we briefly review the concept
of roles as it is used in sociology and then survey attempts by
computer scientists to integrate those concepts in AOP languages
and frameworks. 
%}}} introduction
%{{{ background
\section{Background}

Compared to OOP (object oriented programming), AOP (agent oriented
programming) is done at a higher level of abstraction. The state of
an agent is a set of beliefs, which are described in a logic-like
language. The actions of the agent are dictated by declarative rules.
In AOP style it is easy to simulate societies and to code distributed
algorithms.

\todo{Briefly explain concepts related to agents that are used later,
  \emph{except} roles and organisations. Use examples.}
\todo{Talk about FIPA specs.}
\todo{Talk about JADE.}
\todo{Talk about other AOP languages and frameworks that do \emph{not}
  have roles, but are referred to later, perhaps because they were
  extended with roles.}

The Java Agent DEvelopment framework complies with the FIPA
specifications and is implemented in Java. Several systems
including roles are built on top of it. \todo{Say exactly which
are those systems.}
%}}} background
%{{{ roles in sociology
\section{Roles in Sociology}

Roles are sets of connected behaviours, rights, and obligations
that correspond to the social position of a person (agent) in a
group (organisation). This definition is used both by functionalists
(for example, Auguste Comte) and by modern and contemporary
interactionists.
%}}} roles in sociology
%{{{ roles in software engineering
\section{Roles in Software Engineering}
\label{sec:software_engineering}

\todo{Explain the typical relation between agents, roles, and
  organisations.}
\todo{Explain the difference between a (conceptual) model and an
  implementation. Make it clear what is the difference between a
  model and a meta-model.}
AOP (agent oriented programming) is supported by languages and by
frameworks.  \todo{Explain the difference between languages and
frameworks.}


\todo{More intro.}

\todo{For the following, make it clear which are conceptual models and
  which are implementations.}
\todo{Decide when you say `multi-agent system' and when `agent system'.
  It's probably OK to stick to just one.}

\subsection{3APL, 2APL}

-- 3APL \cite{books/sp/map2005/DastaniRM05} is a programming language for implementing cognitive agents that have
beliefs, goals, plans, actions and reasoning rules as mental attitudes. --
informatica Survey \cite{journals/informaticaSI/BordiniBDFGLOPR06}

--There are plans for extending 3APL to include roles: a role is defined
through a set of predicate goals, a set of interaction rules and the norms
associated with the role (obligations and permissions). -- Role-Assignment in
Open Agent Societies (Dastani et al) \cite{conf/atal/DastaniDD03}

--supports the integration of Prolog and Java -- informatica Survey

\subsection{AFAPL2}

extended version of AF-APL \cite{conf/promas/RossCO04}
--The mental state consists of two primary mental attitudes: beliefs and 
commitments. -- informatica Survey

\subsection{AGENT0}

\cite{journals/ai/Shoham93}

\subsection{AgentSpeak}

\cite{conf/maamaw/Rao96}

\subsection{GOAL}

\cite{hindriks2009programmingrationalagents}

\subsection{JACK Agent Language}

\cite{howden2001jack}
-- extension of Java allowing programmers to create plans and belief bases.
Other features: reasoning methods, capabilities, teams of agents -- informatica
Survey 

\subsection{Jason}

\cite{books/sp/map2005/BordiniHV05}
-- extension of AgentSpeak -- informatica Survey

\subsection{J-Moise+}

\cite{journals/ijaose/HubnerSB07}
Jason, AgentSpeak, Moise.

\subsection{MetateM}

\cite{conf/rex/BarringerFGGO89}

\subsection{PLACA}

\cite{}

\subsection{RoleEP}

\cite{ubayashi2002roleep}

\subsection{ROPE}

\cite{conf/coopis/BechtGKM99}

\subsection{Normative multi-agent programming language}

\cite{conf/promas/TinnemeierDM08}

\subsection{SPLAW}

\cite{}

\subsection{SLABSp}

\cite{conf/atal/WangSZ05}

\subsection{Truce}

\cite{conf/coordination/JamisonL99}

\subsection{AALAADIN} 
\label{sec:aalaadin}

AGR \cite{conf/aose/FerberGM03}, AALAADIN's meta-model, is a model for organisation-centered
multi-agent systems. AGR is based on three main concepts: agents,
groups and roles.

An organisation is constituted of agents that manifest a behaviour.
The overall organisation can be partitioned into groups that may
overlap. 

Agents play roles within groups. A group can be viewed as a
context for activities. Two agents can interact only if they are
members of the same group, but usually an agent is member of
several groups. Groups are sets of agents that either have some
similar characteristics or work together toward a common goal.

A role describes an abstract behaviour of agents and represents the
abstraction of a functional position of an agent in a group.

\subsection{BRAIN} 

The BRAIN Framework \cite{conf/coopis/CabriLZ03} supports the development of
interactions in agent-based applications, using the concept of role. It
consists of (1)~an interaction model, (2)~an XML-based notation to express
roles (XRole), and (3)~an interaction infrastructure implementation
(Rolesystem).

A role is a set of capabilities and has an expected behaviour.
Capabilities are activities that an agent can perform. Behaviours
are reactions to incoming events. \todo{What is a `reaction'?
Does a role have one or multiple behaviours? Does a behaviour
prescribe one action for one event or multiple actions for a
combination of events?}

Rolesystem is built on top of Jade. It translates actions of
agents into event notifications to other agents. An event may,
in turn, trigger actions, according to the roles assumed by
the notified agent. In this way, Rolesystem, facilitates the
interaction between agents.

\subsection{Power Jade}

\todo{Does this have a better name than `Jade extension'?}

\todo{This paragraph seems to serve as a good general description of the
relation between agents, roles, and organizations.} Baldoni et al.
\cite{conf/mates/BaldoniBGGT08} extended the Jade framework to offer primitives
for constructing organisations. The underlying conceptual model is based on the
concept of role. Roles are used to structure the organisation, to distribute
responsibilities among agents and as a means of coordination. A role exists
only in the context of an organisation, working as an interface between the
organisation and agents. Roles offer powers to the agents to operate inside the
organisation, and in the same time request a set of requirements from them. The
expected behaviour is described in terms of goals and beliefs attributed to a
role. An agent can play multiple roles in an organisation but not
simultaneously, only one role at a time can be active.  \todo{addDROPS}
\todo{Explain the previous TODO.}

\subsection{JadeOrgs}

JadeOrgs \cite{madrigal2009adding} extends Jade. Unlike a Jade agent, a
JadeOrgs agent manages roles and organisation membership. A JadeOrgs
organisation is a Jade agent that manages a group of JadeOrg agents.
\todo{`Manages' sounds a bit like `leads', when what I mean is that it
bookkeeps membership. Rephrase.} Roles consist of (1)~required behaviours,
(2)~required ontologies, and (3)~templates for messages. \todo{Huh? Why would a
role have templates for messages? Explain.}

\subsection{Moise+} 

The Moise+ \cite{conf/sbia/HubnerSB02} organisational model is the first one to
join roles with plans. It consists of three main dimensions: structural,
functional and deontic. Roles represent the structural, plans are the
functional, and the relations between roles and plans represent the deontic
dimension. \todo{Define the word `deontic'.}

\todo{Decide on some shape for the following three paragraphs and use the same
shape for all three.}

The structural dimension has three levels. The individual level is represented
by roles, which are viewed as sets of constraints that agents ought to follow.
The social level is represented by links, which are relations between roles,
and constrain the agents' behaviour in relation to other agents. The collective
level is represented by groups, which introduce compatibility constraints
between roles, which specify what roles an agent is allowed to assume,
depending on the roles that it is currently playing. \todo{Perhaps this
paragraph could be made to flow better.}

The functional dimension is based on the concepts of missions (sets of coherent
goals that an agent can commit to), global plans (goals in a structure) and
social scheme (the goal decomposition tree). 

The deontic dimension defines at the individual level the set of obligations
and permissions of a role on a mission.

Saci-Moise+, Jason-Moise+ and Moise+JavaAPI frameworks are all implementations
of the Moise+ meta-model. \todo{More detail.}

\subsection{MadKit} 

MadKit \cite{conf/agents/GutknechtF00a} is a multi-agent platform written in Java. It is build upon the AGR
organisational model (Section~\ref{sec:aalaadin}).

MadKit supports heterogeneity in agent architectures and communication
languages. Agents can be programmed in Java, Scheme, Jess and BeanShell.

\subsection{GAIA}

Gaia \cite{journals/aamas/WooldridgeJK00} is a two phased methodology for analisyng and designing multiagent
systems.

The analysis phase aims to collect and organise specifications. This includes
identification of organisational goals, expected global behaviour, and
decomposition in sub-organisations. Its output consists of a model of the
environment, a preliminary model of roles, a preliminary model of interactions,
and organisational rules.

The desingn phase is split in two: the architectural design and the detailed
design. The architectural design includes the definition of the organisational
structure in terms of its topology and control regime, as well as the
completion of the preliminary role and interaction models. The detailed design
phase covers the definition of the agent model based on the role model and the
definition of the services model, where services correspond to blocks of
activity required to realise agent's roles.

\cite{BlakeG05}
\cite{conf/aose/WooldridgeC00}

\subsection{MaSE}

MaSE methodology \cite{deloach2004mase} is an organisational approach to
analysis and design of MAS. It also supports the implementation through the
associated AgentTool development system. Both MaSe and AgentTool are independent of any particular agent architecture, programming language or communication framework.

The Analysis phase define system goals from a set of requirements and then define the roles necessary to meet the goals. A role consists of a number of tasks, where a task corresponds to a goal. Each task has its own thread of control, but may communicate with each other.

The Design phase assigns roles to agent classes, identify and constructs conversations (coordination protocols between two agents), define the internal agent architecture and the final system structure. 

\subsection{OMNI}

OMNI, the Organisational Model for Normative Institutions \cite{journals/aamas/Vazquez-SalcedaDD05} consists of three components: Normative, Organisational and Ontological dimensions. All of them can be considered at three different levels of abstraction: abstract, concrete and implementation level.

The Normative dimension focus on abstract norms, concrete norms and rules and procedures. The Organisational dimension specifies the social structure through Organisational Models (roles and their interactions), Social Models (the enactment of roles by agents) and Interaction Models (interaction contracts as concrete interaction scenes).The Ontological dimension includes domain ontologies (concrete and procedural) as well as the communication acts (preformatives, protocols, scene scripts) as a form of abstract conversations.

Depending on the domain of application's type the design of multiagent systems with OMNI is guided either by the Organisational or by the Normative dimension.

\subsection{SODA}

SODA \cite{conf/aose/Omicini00} is an agent oriented methodology for the analysis and design of Internet-based systems. It focuses on the social aspects  exploiting coordination models and technologies. The key concepts are tasks, services, society (with a clear separation between individual an social) and environment. SODA doesn't address any intra-agent issues.

During the Analysis phase three models are generated: Role Model,based on tasks (composed of responsibilities, required competences, and resources), roles (consisting of individual tasks, permissions to access resources, and interaction protocols) and groups (formed of social tasks, permissions to access resources, social roles, and interaction rules). Resource Model, based on services as abstract resources representing the functionality provided by the agent environment. Interaction Model, consisting of interaction protocols and interaction rules.

The Design phase results in three other models: Agent Model -- individual and social roles are mapped upon agent classes. Society Model -- groups are mapped onto a society of agents. Environment model -- resources are mapped onto infrastructure classes.

\subsection{Styx}

The Styx methodology \cite{bush2001styx} is targeted to the development of collaborative agent systems with less than 30 roles. Roles are represented as agent classes. Styx methodology covers all of the life-cycle activities: 

Analysis -- focuses on the identification of agent roles and domain concepts and generating Use-Case Maps and Domain Concepts Models; 

Design -- results in Role Responsibility Models, Role Relationship Model (relationships between roles) and Deployment Model (maps roles to agents); 

Implementation -- automatically generates Agent Skeleton in two steps 1) design models are mapped into an abstract agent specification language and 2) the agent specification language is mapped to a particular implementation language, toolkit or platform. It leaves the programmers to focus on the application specific aspects.

Maintenance -- updating the skeleton implementation with the application specific code in place.

Other interesting features of the Styx methodology include specifying of agent message contents as UML class diagrams and reusing FIPA interaction protocols.

\subsection{Tropos}

Tropos \cite{journals/informaticaSI/SusiPMG05} is an agent oriented software development methodology founded on the key concepts of actor, goal, plan, resource, social dependency, and the i* modeling framework. Relationships between actors are depicted in the actor diagrams, while the relationships between goals are represented in goal diagrams. Agent, role, and position are viewed as specialisations of the concept of actor.

An extended version of the metamodel includes the concepts of ownership, delegation, trust, and provisioning. The owner has full authority concerning the access and disposition of a resource, goal or plan.

Tropos supports four phases of software development: Early requirements analysis (understanding the organisational context), Late requirements analysis (functional and non-functional requirements), Architectural design (goal structure), and Detailed design (specification of communication and behaviour of agents). 

A number of tools increase Tropos' usability: TAOM4e -- a modelling environment integrated into Eclipse, T-Tool -- verification of requirements specification through model checking, GR-Tool -- for reasoning on the goal analysis structures.

%}}} roles in software engineering
%{{{ conclusion
\section{Conclusion}
%}}} conclusion

\bibliographystyle{plain}
\bibliography{survey,bib}

\end{document}
