\documentclass{article}
\title{Roles in Agent Oriented Programming\\\normalsize{a survey}}
\author{Claudia Grigore, Rem Collier, Radu Grigore}

\begin{document}
\maketitle

\section{Introduction}

Compared to OOP (object oriented programming), AOP (agent oriented
programming) is done at a higher level of abstraction. The state of
an agent is a set of beliefs, which are described in a logic-like
language. The actions of the agent are dictated by declarative rules.
In AOP style it is easy to simulate societies and to code distributed
algorithms.

Clearly, AOP is inspired by sociology. Recently there has been
much interest in importing other concepts from sociology, roles
and organisations, in order to make it easier to write large scale
programs. In this paper we briefly review the concept of roles as it
is used in sociology and then survey attempts by computer scientists
to integrate those concepts in AOP languages and frameworks. We assume
that the reader is familiar with the basic concepts of AOP (see, for
example, \cite{todo}).

\section{Roles in Sociology}

Roles are sets of connected behaviours, rights, and obligations
that correspond to the social position of a person (agent) in a
group (organisation). This definition is used both by functionalists
(for example, Auguste Comte) and by modern and contemporary
interactionists.


\section{Roles in Software Engineering}

AOP (agent oriented programming) is supported by languages and by
frameworks.

\subsection{Roles in AOP languages}

\subsubsection{AGENT0}
\subsubsection{AFAPL2}
\subsubsection{PLACA}
\subsubsection{SPLAW}
\subsubsection{SLABSp}
\subsubsection{Jason}
\subsubsection{3APL, 2APL}
\subsubsection{GOAL}
\subsubsection{MetateM}
\subsubsection{J-Moise+}
\subsubsection{Normative multi-agent programming language}

\subsection{Roles and Organisations in the AOSE Meta-Models}

\subsubsection{AALAADIN} The AALAADIN's meta-model, AGR, is described
as being a model for organisation centered multiagent systems. AGR is
based on three main concepts: agents, groups and roles.

An organisation is constituted of agents that manifest a behaviour.
The overall organisation can be partitioned into groups that may
overlap. 

Agents play roles within groups. A group can be viewed as a context
for activities. Two agents can interact only if they are member of
the same group, but usually an agent is member of several groups.
Groups are sets of agents that are sharing common characteristics, and
represent either sets of similar agents or function based systems.

A role describes an abstract behaviour of agents and represents the
abstraction of a functional position of an agent in a group.

\subsubsection{BRAIN} The BRAIN Framework is intended to support
the development of interactions in agent-based appllications, using
the concept of role. It is constituted of three components: the
interaction model, XRole -- an XML-based notation to express roles,
and Rolesystem -- an interaction infrastructure implementing the
proposed model.

A role is modeled as a set of capabilities -- activities that can be
performed, and an expected behaviour -- reactions to incoming events.
Roles are temporary, generic and related to contexts.

It is considered to be an interaction between two agents when an
agent performs an action, from the set of actions associated to its
role, and that action is translated into an event that is notified to
another agent which is expected to react conforming to its role.

The underlying interaction infrastructure provides the translation
from actions to events. In this particular case, the Rolesystem
implementation is written in Java, and it is built on top of the Jade
Platform.

\subsubsection{JADE} The Java Agent DEvelopment framework is
fully implemented in Java language, and complies with the FIPA
specifications. It has been extended to include roles and
organisations as follows:

Baldoni et al. extended the Jade framework to offer primitives
for constructing organisations. The underlying conceptual model
is based on the concept of role. Roles are used to structure the
organisation, to distribute responsibilities among agents and as
a means of coordination. A role exists only in the context of an
organisation, working as an interface between the organisation and
agents. Roles offer powers to the agents to operate inside the
organisation, and in the same time request a set of requirements
from them. The expected behaviour is described in terms of goals and
beliefs attributed to a role. An agent can play multiple roles in an
organisation but not simultaneously, only one role at a time can be
active. TODO: addDROPS

JadeOrgs is an extension to JADE that introduces agent organisations
and roles. The JadeOrgs metamodel is centered around three concepts:
JadeOrgs Agent, Organisations, and Role. JadeOrgs Agent is a
specialisation of Agent that manages organisation membership
information and roles. The Organisation is also a specialisation
of Agent with added functionality related to associated roles,
registration and deregistration of members. Roles integrate required
behaviours and ontologies, along with message templates for the
messages sent and received.

\subsubsection{Moise+} The Moise+ organisational model is the first
one to join roles with plans. It consists of three main dimensions:
structural, functional and deontic. Roles represent the structural,
plans are the functional, and the relations between roles and plans
represent the deontic dimension.

The structural dimension has three levels. The individual level is
represented by the roles, that are viewed as sets of constraints
that agents ought to follow. The social level is represented by the
links, which are relations between the roles, and constrain the
agents behaviour in relation to other agents. And the colective level
is represented by the groups, which introduce the compatibility
constraints between roles, that specifies what roles an agent is
allowed to assume depending on the roles that it is currently playing.

The functional dimension is based on the concepts of missions (sets of
coherent goals that an agent can commit to), global plans (goals in a
structure) and Social Scheme (the goal decomposition tree).

The deontic dimension defines at the individual level the set of
obligations and permissions of a role on a mission.

Saci-Moise+, Jason-Moise+ and Moise+JavaAPI frameworks are all
implementations of the Moise+ meta-model.

\subsubsection{MadKit} MadKit is a multiagent platform written in
Java. It is build upon the AGR organisational model, which was
described earlier.

MadKit supports heterogenity in agent architectures and communication
languages. Agents can be programmed in Java, Scheme, Jess and
BeanShell.


\subsubsection{GAIA}
\subsubsection{MaSE}
\subsubsection{OMNI}
\subsubsection{SODA}
\subsubsection{Styx}
\subsubsection{Tropos}

\section{Conclusion}

\bibliographystyle{plain}
\bibliography{survey}

\end{document}
