\documentclass{article} 
\usepackage{hyperref}
\usepackage{xcolor}

\newcommand{\todo}[1]{[\textcolor{red}{TODO}: #1]}

\title{Progress Report 2010: grigore\_claudia\_2008\_interim\_report} 
\author{Claudia Grigore}

\begin{document} \maketitle

\section{Objectives}
Agent Oriented Programming (AOP) is a relatively new programming paradigm which
adopts a social metaphor for the design and implementation of software systems.
Specifically, software systems are viewed as communities of software entities,
known as agents, that interact with one another in order to solve problems that
are beyond their individual capabilities. It is widely accepted that this
approach is well suited to problem domains in which there is no global system
control, data is decentralised and computation is asynchronous.

Organisations are increasingly seen as an important concept in agent oriented
design. However, little work has been done on applying organisational concepts
to AOP languages. For example, integrating the concept of roles offers
significant advantages. They are a natural metaphor for describing the overall
system behaviour, and they increase the adaptability and flexibility of agent
systems by offering the appropriate level of granularity. Roles define a set of
related behaviours, encapsulating them realises a separation of concerns and
promotes information hiding.

The primary hypothesis of my research is that the integration of roles, based
on sociological Role Theory, will improve the readability and usability of AOP
languages. 

The objectives of this research for the last twelve months can be grouped in
two main categories. The first one includes writing a survey of roles in agent
oriented programming languages, and writing the PhD proposal required in UCD
for the transfer to the second stage of the PhD. 

The second category of objectives is modifying the AFAPL2 language to support
the Moise+ methodology. This includes allowing agents to change their roles at
run-time, and to reason abut their roles and about their interrelations.

\section{Methodology}
In order to achieve the categories of objectives I have set, I have adopted the
following approach:

\begin{itemize}
\item attending courses, seminaries, and summer schools on various agent topics
to broad my view and understanding of the area.

\item reading and summarising papers on agents, roles, and organisations, for a
deeper understanding of the specific problem of integrating organisational
concepts into AOP languages.

\item solving concrete problems to attain a better understanding of MAS
toolkits and AOP languages.

\end{itemize}

\section{Results}

\begin{itemize}
\item attended courses
  \begin{itemize}
  \item Principles of Programming Languages
  \item Object Oriented Design
  \item Social Networks and Agent-based Simulation
  \item Research Skills
  \item Research Seminaries
  \end{itemize}
\item attended the EASSS 2009 summer school in Torino
\item explored the use of Social Network Analysis techniques in agent systems
\item 
  researched the current state of the art in the area of organisations in MAS,
  summarised in the survey
  \begin{itemize}
  \item theoretical models of organisations
    \begin{itemize}
    \item Moise+
    \item AALADIN
    \item BRAIN
    \item GAIA
    \item TROPOS
    \item OMNI
    \end{itemize}
  \item implementations
    \begin{itemize}
    \item MADKIT
    \item S-Moise+
    \item Jade
    \end{itemize}
  \end{itemize}
\item understanding MAS toolkits by solving problems
  \begin{itemize}
  \item AgentFactory
  \item S-Moise
  \item Jade
  \end{itemize}
\item presentation on MAS in UCD's Research Seminar Series
\end{itemize}


\section{Significant Difficulties Encountered}

\todo{\emph{Research} difficulties.}

Course work and demonstrating was more time consuming, delay of 2
months.

Maternity leave of 3 months. 

Baby sick: chicken pox, bronchitis, then urinary tract infection -- 2 months
delay.

\section{Proposed Work for the Next 12 Months}

PhD transfer report due end of June.

Specify and implement a prototype of a role-based AOP language.

\{??integrate AFAPL2 with S-Moise+??\}

\section{Publications}

\begin{itemize}
\item Survey. \todo{target}
\item Moise+ in AgentFactory. \todo{target}
\end{itemize}

\end{document}
