%{{{ latex prelude
\documentclass{article}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{hyperref}

\def\fb#1{\textbf{#1}}

\newcommand*\R[1]{\textcolor{red}{#1}} % used for refs that might change
\newcommand{\todo}[1]{[\textcolor{green}{TODO}: #1]}

\newenvironment{them}%
  {\bigskip\noindent\begingroup\color{blue}$\blacktriangleright$\enspace}%
  {\endgroup\par}

\title{Correction Report}
\author{Claudia Grigore}
% latex prelude }}}

\begin{document}
\maketitle

%{{{ introduction
\section{Introduction}
The original thesis had undergone drastic changes, starting with the title: the chapter structure was altered, and the size has doubled. Many chapters were added, including one on Implementation, and one on Multiparty session types; a new approach to evaluation was devised, including a user study; all objectives were mapped to the AF-Raf concepts, as well as to the thesis content. The AF-Raf agent architecture is described both textually and in a graphical format. Technical errors were inspected and corrected. Syntax and semantics are correct and consistent.

I have addressed all 116 issues raised in the PhD report carefully.
Details are presented in the following section.

%}}} introduction
%{{{ corrections
\section{Corrections and Clarifications}

\begin{them}
-- Title

The title of the thesis is very general and refers neither to the introduction
of roles, sessions and type checking into the agent oriented programming
paradigm, nor to the AF-Raf programming language which has been actually
extended. A more precise title is required.
\end{them}
Fixed. The title of the thesis is now ``AF-Raf, a New Agent Oriented Programming
Language with Roles, Sessions and Type Checking''.

\begin{them}
-- Structure of the thesis

A ``Background'' chapter is missing; many background notions are scattered
throughout the thesis, whereas they should be included in a chapter, just after
the Introduction, as customary.
\end{them}
Fixed. All background notions are now incorporated into the Background (\R{Part
I} of the thesis).

\begin{them}
One chapter devoted to the ``Implementation'' of AF-Raf (with link to an online
prototype) and one devoted to ``Experiments'' should be added before the
Conclusions.
\end{them}
Fixed. I've added a chapter devoted to AF-Raf implementation \R{Chapter 14}.
Also, I've added three chapters (\R{15}, \R{16}, \R{17}) devoted to experiments
and evaluation which constitute \R{Part III} of the thesis.


\begin{them}
The thesis has four main objectives: 1a: integration of roles as first class
entities into the AOP paradigm in general, and into AF-Raf in particular; 2a:
exploration of the analogy between AOP and functional languages; 2b:
integration of multiparty session types into the AOP paradigm in general, and
into AF-Raf in particular; 2c: integration of type checking capabilities into
the AOP paradigm in general, and into AF-Raf in particular.  The structure of
the thesis should reflect these four objectives and should allow the reader to
clearly identify in which point of the thesis each objective is dealt with. In
particular, for objectives 1a, 2b, 2c a high level description of the design of
how the objective can be achieved within a programming language in general, a
detailed description of how it was achieved in AF-Raf, a discussion of the
implementation details, a well-motivated evaluation of the benefits of having
achieved that objective, and a careful comparison with the related work should
be provided.  One way to structure the thesis in such a way that each objective
is suitably covered might be to devote a complete chapter (or more) to each
objective. Another way might be to keep the structure of the thesis close to
the current one, but clarifying with proper sections and subsections which
objective is being dealt with in each chapter.
\end{them}
Fixed. 

\R{Chapter 17}, Results and Evaluation, clarifies where and how each objective
was achieved.

Also, I've added in \R{Section 1.2} an extra subsection to define how the
achievement of objectives is evaluated, as well as a new  section \R{(1.3,
Objective Mapping)} to clarify which parts of the thesis deal with what
objective.

\begin{them}
The format and graphical layout of the thesis should be homogeneous and should
help the reader in easily finding the titles of sections and subsections.
\end{them}
Fixed. The table of contents lists all the titles of sections and subsections

\begin{them}
Examples should not be provided inline with the text.
\end{them}

I have checked that the examples throughout the thesis conform to normal
standards of mathematical writing: very small examples are interspersed with
text that presents informal high level ideas, more involved examples have their
own environment, and large, significant examples have their own figure.

\begin{them}
-- Evaluation and comparison with the related work

The evaluation of the proposed approach is not strongly supported; the
comparison with the related work is somehow limited. Whatever the structure of
the thesis, a well motivated evaluation and a deep comparison with the related
work should be provided for each of the four objectives of the thesis.
\end{them}
Fixed. Evaluation had been deepened with the aid of a custom made survey. The
results are presented in \R{Chapter 16}. Also, comparison with related work has
been improved, see \R{Chapter 11}.

\begin{them}
-- High level description, syntax and semantics of the proposed AF-Raf language

The high level description of AF-Raf might should include the description of
the AF-Raf architecture of one single agent (which are the AF-Raf agent's
components? For example, a rule library, a belief library, a session library, a
set of execution stacks...) and of the control flow, or interpreter, working on
them. If some AF-Raf components are defined at the MAS level rather than at the
agent level, the AF-Raf MAS architecture and control flow should be given as
well.
\end{them}
Fixed.

I have included a graphical representation of an AF-Raf agent in \R{Figure~12.5
}. It depicts all AF-Raf agent components as well as the control flow. The
AF-Raf architecture of an agent is described in \R{Section~12.1}. AF-Raf agent
components: rule base, belief base, roles, sessions, type checking are all
described there. Detailed explanations regarding the sensors and actions are
given in the same section starting with \R{page 66}. The MAS architecture and
control flow, including type checking is depicted in \R{Figure~12.5}. The
interpreter is described at the beginning of \R{Section~13.4} and in
\R{Chapter~14}.

\begin{them}
An explanation of how the basic features of agents (autonomy, situatedness,
proactivity, reactivity, sociality) are supported by the language should be
provided. 
\end{them}
Fixed. I've added \R{Section 2.1} to define the AF-Raf agency.

\begin{them}
The relationships between AF-Raf agents and the environment where
they live should be explained: how are the agents' sensors and effectors
implemented? 
\end{them}
I have improved the explanation in \R{section 11.1}.

\begin{them}
Syntax and semantics should be correct and consistent throughout the entire
thesis. In the detailed comments below a subset of the technical errors found
are highlighted and discussed, but the revision made by the candidate should go
beyond the specific points raised by the reviewers.
\end{them}
Fixed. Syntax and semantics are correct and consistent.

\begin{them}
-- Fluency

The thesis requires a revision by a native English speaker, as many typos and
involved sentences were found which should be removed in the final version.
\end{them}
Fixed. Thesis reviewed by Florence Coetzer, BA in Classics.

\begin{them}
Detailed Comments

-- Chapter 1

Page 1:
The usual references to the definition of agents and MASs are the pre-2000
papers by Wooldridge and Jennings (KES 1995) or by Jennings, Sycara and
Wooldridge (JAAMAS 1998): reference 46 does not seem appropriate.
\end{them}
Fixed. I've included references to Wooldridge and Jennings (KES 1995) paper.
I've also used this paper when defining AF-Raf agency, in \R{Section 2.1}.

\begin{them}
Page 2:
Stating that the set of standards by FIPA is ``nascent'' is debatable, as it is
very old.
\end{them}
Fixed.

\begin{them}
Reference to AgentFactory is very old, something more recent should be used
instead (if it exixts).
\end{them}
Fixed.

\begin{them}
ZEUS is a dead project, maybe it is not worth citing.
\end{them}
Fixed.

\begin{them}
The methodologies cited in the middle of the page are almost old: additional
more recent references should be looked for and added.
\end{them}
Fixed. I've added newer methodologies.

\begin{them}
``This relatively new paradigm'' is debatable, as AOP is more than 20 years old.
\end{them}
Fixed.

\begin{them}
Page 3:
Reference 47 does not meet the standard for bibliographic items.
\end{them}
Fixed.

\begin{them}
References in page 3 are all very old (for example [32]: there are more recent
proposals for metamodels that might be added); this section of the thesis
should be revised in order to be more up-todate.
\end{them}
Fixed.

\begin{them}
Page 5:
The statement that the analogy between AOP and Functional Programming was never
explored is not correct; although it was a preliminary work, the paper by
Solimando and Traverso published in DALT2012 dealt with this issue.
\end{them}
Fixed. Added reference to their work, which is preceded by my publications.

\begin{them}
Page 6:
The sentence ``and design the first AOP language to introduce type checking'' is
not clear,
\end{them}
Fixed.

\begin{them}
and the third objective of the thesis should be better analyzed in
the thesis.
\end{them}
\R{Chapter 8} provides the background for types in AF-Raf. \R{Section~12.1}
gives an informal overview and a graphical representation of type checking in
\R{Figure~12.5}. A detailed explanation of types and type checking in AF-Raf is
conveyed in \R{Section~12.3}. Formal definitions of AF-Raf types are provided
in \R{Section~13.2.2}, along with clarifying examples. Types as arguments in
operations are described in \R{Section~13.3}. For (dynamic) type
checking operational semantics see \R{Section~13.4} from \R{page 142} onwards.
Also, a representative part of the type checker implementation is now in
\R{Apendix~C.3}. An evaluation of the results is summarised in \R{Chapter~17}.

\begin{them}
-- Chapter 2
This chapter should be entitled ``Background'' and all the background knowledge
needed to understand the thesis' contribution should be moved here. The
description of AgentFactory, the BDI model and AgentSpeak(L), SimpAL, JADE, the
global and session types theories and their projection, ALPHA, etc, should be
all introduced here at a level of detail which makes the thesis self-contained
(for a computer scientist).
\end{them}
Fixed. All the background knowledge is now moved to Background, \R{Part I} of
the thesis

\begin{them}
Important topics to introduce in the Background section are agent communication
principles and languages. Asynchronous message passing using high level ACL
like FIPA ACL and KQML should be introduced here in order to better motivate
the ``Functions as Messages'' analogy in page 47, putting it into the right
context.
\end{them}
Fixed. Added \R{section 5.1} on communication.

\begin{them}
In a similar way, what are ontologies, how they are used in JADE and in other
agent programming frameworks, and the usefulness of having an ontology to
support interoperability among heterogeneous agents should be introduced here,
in order to contextualize the statements in page 48.
\end{them}
Fixed. Added \R{section 5.2} on ontologies.

\begin{them}
Page 8:
Again, the ``relatively'' new paradigm is debatable.
\end{them}
Fixed. 

\begin{them}
Page 11:
What does it mean that roles are negotiated on a constant basis?
\end{them}

I believe that what Biddle means when he says that ``roles are negotiated on
a constant basis'' is clear: People's roles throughout life are not fixed,
and the way they change is through negotiation, a process that that never
stops.


\begin{them}
Page 12:
The idea of role oriented programming should be better detailed and explained,
as it is closely connected with objective 1a of this thesis. Powerjava is cited
as both an extension of OOP, and as an extension of AOP: a better understanding
and synthesis of this language is needed. ROPE dates back to 1999; more recent
proposals should be mentioned.
\end{them}
Fixed. Described Object Teams and PowerJava in more detail \R{pages 21-22}.

\begin{them}
Page 14:
Gaia, MASE, AALAADDIN are AOSE methodologies, whereas they are described in the
thesis as ``organizational frameworks''; this point should be better expanded.
\end{them}
The section was updated to provide more accurate and detailed data regarding
the methodologies and software frameworks that promote organisation oriented
programming.

\begin{them}
The sentence ``there is no full support at the language level for the
organization of agents'' is not cleas, as it refers to AOSE methodologies and
not to AOP languages. Also, at least Gaia and MASE include the notions of role
and protocol, so they do provide a support to these concepts. A recent
extension to MASE, o-MASE (JAOSE 2010), integrates organizational concepts. It
might be interesting for the thesis.
\end{them}
This confusing sentence was removed. The new updated section explains better
the need for support at the programming language level. 

\begin{them}
Page 20:
Table 2.1 should be improved and integrated with other features (rows) and
other related proposals (columns) in order to provide a complete comparison
between AF-Raf and other languages.
\end{them}
Fixed. Table improved.

\begin{them}
-- Chapter 3
This chapter should provide a gentle introduction to AF-Raf; it should start
with 1. a clear description of the AF-Raf architecture based on a graphical
representation of the architecture's modules and sub-modules and
\end{them}
Fixed. The AF-Raf architecture is clearly described. The graphical
representation of the architecture is provided in \R{Figure~12.2}

\begin{them}
2. a clear description of the control flow among the architecture's component
described at least in a graphical way and in structured natural language; maybe
a high level pseudo-code could also help as done for example in

\url{http://www.upv.es/sma/teoria/teoria_ag/bdi\%20agents\%20from\%20theoryRao.pdf},
\end{them}
Fixed. The graphical representation of the control flow \R{Figure~12.2} is
accompanied by a textual illustration.

\begin{them}
Section 5.
In the architecture, the relationships between the modules and the thesis'
objectives should be made explicit, as it would be natural to find a module for
supporting each of the objectives 1a, 2b, 2c.
\end{them}
Fixed. Added brief description to link AF-Raf components with the objectives
(1a), (2b), (2c). Also, the relationship between AF-Raf components and
objectives is described in  more detail in \R{Chapter 17}.

\begin{them}
Short and easy to understand fragments of code could be used to exemplify each
element of the language: examples for beliefs, rules, roles, messages, actions,
(software) sensors, (software) effectors, sessions, and so on, should be
provided.
\end{them}
All elements of the language have exemplifying code. 

\begin{them}
Page 23:
Figure 3.1, 5.2 and 5.3 all show the code of an agent but they look very
different from one another, making the reader confused. The caption of each
example and code fragment in the thesis should clearly state which feature of
the language is going to be shown. All the examples should be consistent. If
examples showing the same features are different, their captions should briefly
explain why.
\end{them}
The reader is referred to \R{Section~13.3} (Syntactic Sugar) where this issue
is explained in a formal way and to \R{Section~15.1} (the Calculator) where
this issue is exemplified.

In short, Figure~3.1 (now \R{12.1}) presents an agent code written only with
rules, whereas Figure~5.2 (now \R{15.6}) and Figure~5.3 (now \R{15.7}) present
agent code written using operations (syntactic sugar).

\begin{them}
In the second line of Figure 3.1 a conjunction is used in the rule head, but in
page 31 only queries without conjunctions are shown and the use of ``\&'' as
``and'' is not explained.
\end{them}
Fixed.
Conjunctive queries are defined in \R{(12.30)} in \R{Section~12.2}.


\begin{them}
Page 24:
Figure 3.2 is useless as it is a sub-fragment of Figure 3.1
\end{them}
Fixed. 

\begin{them}
Which is the relationship between the sentence ``executing an action means
executing a piece of associated Java code'' and the situatedness of agents? Are
all these Java actions a form of effectors operating in the environment? How
does sensing take place?
\end{them}

This is now explained in \R{Section~12.1} and \R{Chapter~14}.

Briefly, both actions and sensors are implemented in Java,
  and they both have access to the AgentFactory infrastructure.
Actions are invoked explicitly from AF-Raf code.
Sensors are executed by the AF-Raf interpreter at each time step,
  and they typically have the side effect of updating the belief base.
In turn, the change of the belief base affects
  which rules fire.


\begin{them}
Page 26:
The sentence ``then provides a list containing entries of the form: message
type....'', referring to sessions, should be associated with a BNF giving the
syntax of a session. From the definition of session given in page 26, the
reader understands that a session is a plain sequence of messages. This is
probably false as in page 73 the maybe, repeat, oneof operators - introduced
nowhere in the thesis - are mentioned in the BNF, anyway page 24 gives a
limitative view of sessions (or global types, meaning the global description of
the interaction protocol) as they usually include operators for representing
choices, interleaving, concatenation of sessions, synchronization. 
\end{them}

As explained now in the introductory chapter on multiparty asynchonrous session
types, AF-Raf sessions are only informally similar to global types --- there is
no formal connection, but one can find an informal comparison in
\R{Section~9.4}. In \R{Figure~13.1}, one sees the abstract syntax of sessions;
we remind the examiner that the notation is explained at the beginning of
\R{Section~13.2.2}, and that there are good reasons why it is not EBNF.  The
semantics of AF-Raf sessions appear in the section on semantics.  For example,
see the the rule \R{(13.72)}.

In summary,
  AF-Raf sessions are now introduced informally (\R{12.5}),
  their formal syntax is given (\R{Figure~13.1}),
  and their operational semantics as well (rule \R{(13.72)} and related).


\begin{them}
In the Background section a detailed description of formalisms for global types
and multiparty session types should be provided, together with a comparison of
their expressive power and the expressive power of the formalism supported by
AF-Raf. One of the best papers on the topic is, in the reviewers' opinion,
\url{http://arxiv.org/pdf/1203.0780.pdf}. The candidate should acquire a good
understanding of its content. Each operator of some standard formalism for
global types (or session types, depending on the global vs local perspective)
should be put in relation with the AF-Raf operator offering the same
functionality. When the reader reaches page 26, the notion of global and
session types should be already clear, and their integration into AF-Raf should
be given in a correct and complete way.
\end{them}
Fixed. See \R{Chapter~9}.

\begin{them}
Dynamic and static checks are shortly mentioned in page 26: the topic should be
dealt with in more detail as type checking is one of the thesis' objectives.
\end{them}
I've added a graphical representation of dynamic and static type checking
in \R{Figure~12.5}. More detailed explanation of type checking is given
in \R{Section~12.3.2}. A representative part of the implementation is now
in \R{Appendix~C.3}.

\begin{them}
Page 27:
Section 3.1.1. should be moved in the Background.
\end{them}
Fixed. 

\begin{them}
Page 29:
The sentence ``an update corresponds to a change in the environment'' seems not
appropriate for a section dealing with the Belief Base. A section dealing with
the representation and management of the Environment in AF-Raf should be added.
\end{them}
I've updated \R{Section~12.1} to include more details about the way the
environment is represented and managed. In short, the environment is
represented in the form of beliefs and it is managed accordingly. Given this
piece of information, I believe that the sentence in question is a legitimate
clarification.

\begin{them}
Page 30:
For sake of clarity, syntax and semantics should be kept separate; in page 30
the focus seems to be on the syntax; the sentence ``funtion symbols can
represent both values and typing information'' deals with the semantics and
should be moved in a more suitable place.
\end{them}

I agree that there should be a formal presentation of the syntax
  and a formal presentation of the semantics,
  which should be kept separate.
However, the thesis also should present examples
  that illustrate both syntax and semantics.
I believe it is completely standard to give a figure
  with the syntax of a small language,
and then to comment in text informally about what each construct
  is supposed to do,
  \emph{before} (and if) the formal semantics are given.
For an example of this,
  see Section~2.1 of Honda et al, J.~ACM~2016.


\begin{them}
Page 31:
The sentence ``a query is a sentence in some logic'' is too general, as this
section should provide the AF-Raf syntax, not a general introduction of what
queries are. The candidate should state in which logic AF-Raf queries are
expressed, which operators can appear in a query, and should provide examples
demonstrating many different forms of queries, not just ``atomic'' ones.
\end{them}
Fixed.
% The first sentence is just an introduction.
The logic is a subset of quantifier free first order logic,
  and is given by \R{(12.7--14)}.
Queries are defined in~\R{(12.28--30)}.

I would note that the chapter on AF-Raf concepts
  comes before the language definition,
  as a gentle, informal introduction.
Queries, including composite ones,
  were already defined in the language definition part,
  in what is now~\R{(13.48--51)}.

\begin{them}
Page 32:
Ground terms are defined in page 36, they should be defined here.
\end{them}
Fixed.

\begin{them}
Is parallel substitution really necessary? Since beliefs are ground, maybe it
is not. Please check.
\end{them}

Parallel substitution is indeed not necessary, but covenient.
The new \R{Chapter 9},
  which reviews multiparty session types,
  also uses parallel substitution,
    to be consistent with related literature.


\begin{them}
Page 34:
The candidate should introduce types only after the syntax of each element of
the language has been fully explained. For example, at the end of the page, the
candidate refers to the type of messages, but messages were not introduced
before.
\end{them}

\R{Sections 12.4~and~12.5} depend on \R{Section~12.3},
  so it cannot be delayed.
I would note that \R{Chapter~12} is meant as a gentle, informal
  introduction to AF-Raf concepts.
The language definition is in \R{Chapter~13}.
This is now mentioned in the beginning of \R{Chapter~12}.

\begin{them}
Given the definitions 3.1 and 3.2 in page 30, the only type for a belief is
``string''. Since each belief has type string, it is difficult to understand this
section about typing.
\end{them}
According to those definitions,
  beliefs are not strings but terms.
(These terms could be represented by strings in a computer,
  which is what those definitions suggest.
But that does mean that the \emph{sort} of each term is string.
Sorts are introduced in the next section.)

\begin{them}
The ``Bad scenario'' description is too involved: ``tau''s should be substituted
with concrete messages in order to make this example clearer.
\end{them}
Fixed.

\begin{them}
Page 35:
Section 3.3.1 deals with semantics; it should be better to state when the
discussion about the syntax has completed and when the thesis moves into
semantics issues.
\end{them}
Fixed.
This section,
  which recalls the standard notion of multi-sorted logic,
  is moved in the background part (\R{Part~I}).

I would note, though,
  that the section discusses the syntax of multi-sorted logic,
  not the semantics.
Introducing a few model theoretic notions would, of course, be interesting,
  but does not seem necessary for understanding the rest of the thesis.

\begin{them}
Page 36:
Remark 2: ground terms are trees of strings; what are nonground terms? It
should be also noted that this remark deals with syntax inside a section
dealing with semantics.
\end{them}
This remark (which is now \R{Remark~1}) has been rewritten in less laconic
form. I hope it is now clear what it means.

It should be noted, however, that this section is \emph{not} about semantics.
It gives a \emph{syntax} for multi-sorted predicate logic,
  as a support for introducing the standard notion of \emph{sorts}.
No structures are defined, and no semantic interpretation ($\models$) is given,
  because it does not seem strictly necessary for understanding AF-Raf.


\begin{them}
Sentence ``Typically, in mathematical logic,....'' seems a complex way to define
predicates and formulae, which might be given for granted as they belong to the
background of any computer scientist.
\end{them}
Computer scientists often do not take the stratified view
  that is traditional in mathematical logic.
The point of the paragraph is exactly that computer scientists
  might get the wrong idea if they rely on their background.
For example, SMT-LIB has an {\bf ite} construct
  that perverts the view of traditional logicians,
    who do not accept formulas below terms.
AF-Raf follows this traditional strict separation between formulas and terms.
I added a reference to the classic textbook by Shoenfield.

\begin{them}
Page 37:
Remark 3 highlights a problem: which is the candidate's solution to it?
\end{them}

The candidate's solution is presented in the first sentence of that paragraph:
``The definitions given here are in-between what applied scientists tend to
prefer and what pure logicians tend to prefer''.

\begin{them}
Page 38:
Is Proposition 1 an original outcome of this thesis? If yes, it would be worth
emphasising that result. If not, references should be provided.
\end{them}
Proposition 1 is an original conclusion based on the differentiation between
algebraic data types and sorts. 

\begin{them}
Page 39:
+(2,3) should be +(2(), 3()); is this syntax actually used in AF-Raf agents'
programs?
\end{them}

Fixed. It should, indeed, be +(2(),3()). The user of AF-Raf doesn't have to use
this cumbersome representation.

\begin{them}
Page 41:
``Those operations are: ....'' has been already said in page 40.
\end{them}

Fixed. Definition of algebraic data types was moved to Background.

\begin{them}
In section 3.3.3 it would be interesting to see the actual implementation of
type checking, not a general one.
\end{them}

The actual implementation was added in \R{Appendix~C.3}.

\begin{them}
Page 47:
The analogy ``Functions as messages'' should be better clarified, as it seems not
to fit the common understanding that messages are structured according to the
speech act theory, their content is written in some high level language, maybe
using an ontology to give meaning to terms appearing there, and communication
is asynchronous.
\end{them}
Indeed, AF-Raf messages are not structured according to  the speech act
theory.  We use algebraic data types to define the content of the messages,
roles to define possible message exchanges and sessions to define communication
patterns. For a formal definition of messages see the operational semantics of
messages starting on \R{page 138}.

\begin{them}
The relationships between the role Calculator, CalculatorInt and BinaryCalcInt
should be better explained. A diagram might help.
\end{them}
The relationship between role Calculator and BinaryCalc is explained in
\R{section 12.6 and 15.1.1}. BinaryCalcInt and ExprCalc are agents playing the
roles BinaryCalc and Calculator, this is explained in \R{Section 15.1}.

\begin{them}
Page 48:
``Messages'' do not meet the state of the art format (see FIPA ACL, KQML): a
motivation should be provided for that.
\end{them}

In AF-Raf,
  beliefs and messages have the same format,
  which makes it easy to communicate beliefs.
The format used on the wire is another matter,
  and it could be FIPA ACL\null.
This is a trivial implementation issue,
  which becomes important in practice only if AF-Raf agents
  need to communicate with agents written in other languages.
So far, we have not performed such interaction experiments.

At the moment,
  the standard library of AF-Raf does not contain an implementation of FIPA\null.
There is, however,
  nothing preventing a programmer from implementing a sensor/action pair
  for transmitting messages in a particular format.


\begin{them}
The syntax $f::a->b$ was never introduced before.
\end{them}
This notation is taken from Haskell: see \R{Figure~7.1}.
It is then used when roles are first presented through examples,
  at the beginning of \R{Section~12.4}.
Then, in that section, we find out what the notation means:
``In general, $f::a\to b$ means that \dots''
Finally,
  the text after \R{(13.63)} says how the types $a$~and~$b$
  are used for dynamic type checking.


\begin{them}
The paragraph about the analogy between an ontology for arithmetic expression
and the algebraic data types is debatable, as the need to define an ontology
for arithmetic expressions just to be sure that the content of a message
adheres to their syntax is not so evident. Ontologies are used for other
purposes than giving a syntax of a language or defining the types of
expressions. Dealing with their purposes and usage in the ``Background'' section
should help.
\end{them}
In examiner's own words ``\ldots [message] content is written in some high
level language, maybe using an ontology to give meaning to terms appearing
there''

What this paraghraph says is that using algebraic data types to define the
content of messages that consist of arithmetic expressions is more convenient
than using a JADE ontology.

\begin{them}
Page 49:
``In contrast, the earlier role....'' should be explained before.
\end{them}
I believe that introducing role dependencies after describing simple roles
minimises confusion.

\begin{them}
2APL's PG rules and R-rules have not been defined before.
\end{them}
Fixed.

\begin{them}
Page 50:
The statement ``In summary, the vague and informal intuition...'' is not
supported by evidence in the thesis.
\end{them}
It means that the analogy led to two interesting observations, and it does not
say anything about any results.

\begin{them}
Page 51:
The statement ``such a list is very similar to the global types that describe
multiparty sessions...'' is debatable, as global types represent patterns of
interaction among agents which are much more complex than sequences.
\end{them}

There was a mistake: I did not mean global types, but local types.
Also, I toned down `very similar',
  to make it clear that this sentence is only meant to communicate intuition,
and should not be taken as making any formal judgement.


\begin{them}
Page 52:
The problem of projecting global types to session types has been widely
discussed in the literature (see for example Section 4 of
\url{http://arxiv.org/pdf/1203.0780.pdf}, but also
\url{http://arxiv.org/pdf/1304.1902.pdf} and
\url{http://mrg.doc.ic.ac.uk/publications/the-scribble-protocollanguage/invited.pdf}).
References should be given and a comparison should be drawn.
\end{them}

Fixed.
See \R{Chapter~9},
  in particular \R{Section~9.4}.

\begin{them}
Page 53:
The Calculator Example should start with an explanation of why a calculator is
an agent w.r.t. its autonomy, situatedness, reactivity, proactivity, sociality.
\end{them}
This is now answered in \R{Section~15.1.4}.
In addition, that section was completely rewritten.


\begin{them}
Page 55 and 56:
The code fragments would be more readable if they had inline comments
explaining each line of code; if they become too large, they can be splitted
into more figures.
\end{them}
Fixed.

\begin{them}
Page 58:
The concluding remark ``our notions of roles and sessions overlap in great
amount with the multiparty session types developed in the context of the
Pi-calculus'' should be supported by a more precise comparison of the two
formalisms carried out based on the operators they support, on their
expressiveness (do they define regular languages of sequences over messages?
context free languages? ...), on the definition of ``well-formedness'' (not all
global/session types are well formed), on the assumptions for using them
(determinism, no cover channels, and so on). The comparison with the projection
operator should be carried out as well.
\end{them}

There is no formal connection between AF-Raf sessions and global types.
AF-Raf sessions are inspired by global types in a very high-level,
  informal way.
Nevertheless,
  \R{Chapter~9} now explains session types in some detail,
  and comments at length on the (informal) relationship they have with
    AF-Raf sessions.
The latter are a novel and quite different concept.
An in-depth formal comparison is a non-trivial open problem.


\begin{them}
Page 60:
This section should be moved in the Backgroud; despite the title, the
comparison is missing: a careful comparison should be provided.
\end{them}
Fixed. See \R{Chapter 11}.

\begin{them}
Page 61:
What does sentence ``In AF-Raf sessions provide which can be seen as a kind of
protocol...'' mean?
\end{them}
Fixed.
(Should be ``provide a blueprint which''.)

\begin{them}
-- Chapter 4
This chapter should provide more details on AF-Raf w.r.t. chapter 3, that
should be a gentle and high level introduction to AF-Raf. However, all the
components of the AF-Raf language should be introduced in Chapter 3: in Chapter
4, they should just be discussed in more detail.
\end{them}
Fixed. All components are introduced in \R{Chapter 11}, Main AF-Raf Concepts, and discussed in detail in \R{Chapter 12}.

\begin{them}
Page 66:
A link to the reference implementation of AF-Raf mentioned in the second line
is required
\end{them}
Fixed.

\begin{them}
What is an ``agent design''? 
\end{them}

An agent is an instance of an agent design.
(Just like a Java object is an instance of a Java class.)
This was explained on page 66 (now \R{103}).

\begin{them}
Was the ``AgentDeclaration'' already introduced as a component of the agent's
code?
\end{them}
Yes. See the new definition of agent code in \R{figure 12.7}.

\begin{them}
Page 67:
A standard style for BNF should be used, with ``::='' and ``|'' in the body for
separating the various forms that the body may have; inline comments might be
added to enhance clarity; each definition (left hand side of the ``='') should be
used somewhere and each definition used in the body of a BNF rule (right hand
side of ``=''), should be provided in the grammar. If recursive definitions are
interpreted in the standard inductive way, they should have a base; if they are
interpreted coinductively as they define infinite terms, this should be made
explicit. 
\end{them}
A standard BNF cannot be used.
As explained at the beginning of Section~4.1
  (now \R{Section~13.2.2}),
  we use \emph{two} types of rules, not just one like BNF\null.
The two types of rules can most readily be understood as shorthand
  notation for two Java concepts: composition and inheritance.
To this explanation,
  I now added the one in \R{Section~13.2.1},
  and also the Java code (\R{Appendix~C.2})
    which is generated (automatically) from the \emph{code}
      in \R{Figures 13.1~and~13.2}.


\begin{them}
RoelDeclaration is not used 
\end{them}
RoleDeclaration is the data structure used by the AfRaf interpreter to
represent role declarations.

\begin{them}
Type is defined recursively but it has no base 
\end{them}
Yes, that is correct.
Strong induction also has no base.
Also, see the code for class {\tt Type} in \R{Appendix~C.2}.

\begin{them}
TypeDeclaration is not used 
\end{them}
The TypeDeclaration data structure is used by the AfRaf type checker.

\begin{them}
TypeBranch uses a syntax different from the other rules 
\end{them}
That is inheritance,
  as is explained in the beginning of the section.

\begin{them}
AgentDeclaration is not used 
\end{them}
AgentDeclaration is the data structure used by AF-Raf to represent
  agent designs.

\begin{them}
Function is not defined 
\end{them}
Function is a class defined in the AgentFactory framework.

\begin{them}
QualifiedId is defined recursively but it has no base
\end{them}
Yes, having no base case is correct and intended.
As mentioned, strong induction also has no base case.
See also \R{Appendix C.2}.


\begin{them}
Iformula and IplanStep are not defined
\end{them}
These classes are defined in AgentFactory. A paragraph on \R{page 106}  explains this.
Also, \R{Figure 13.4} presents the grammar for formulas, and \R{Figure~13.5}
presents the grammar for statements.

\begin{them}
Page 70:
Stating that an Array is a Role seems not very close to the common intuition of
what a Role is. In the common understanding of agents' roles, a role may be
``Seller'', ``Buyer'', ``Initiator'', ``Participant'', ``Bidder'' (see FIPA Protocols,
for example) and so on. It is difficult to imagine an autonomous agent playing
the role ``Array''.
\end{them}
Anything that exhibits behaviour that can be used by others can be modeled as a
role. The question is if we chose the right level of granularity. Would it be
better to include that behaviour along with other related ones in an ampler,
more complex role? In this particular situation chosing to model the role
``Array'' is perfectly legitimate, as the purpose of this choice is to give a
very simple and easy to understand example that illustrates the use of type
variables as arguments in a role.

\begin{them}
Page 72:
References to LL(k) grammars and ANTLR should be given.
\end{them}
Fixed.

\begin{them}
Page 77:
``keyword new must be followed by the category of the name....'': what do you
mean by category of the name?
\end{them}
Fixed.

\begin{them}
Sentence ``The core of AF-Raf performs no hidden actions'' should be moved (and
better explained) at the beginning of Chapter 3, when the AF-Raf architecture
and control flow are described.
\end{them}
This sentence referred only to the fact that the received messages have to be
removed explicitly from the belief base. I have made this clear.

\begin{them}
Page 78:
If syntactic sugaring and desugaring are really relevant for the proposed
approach, they should appear in the AF-Raf control flow in the beginning of
Chapter 3 and should be introduced there, leaving the detailed description in
Chapter 4: in general, syntactic sugar does not seem so relevant (it's just
sugar, and as such it could be avoided) so it is not easy to uderstand why it
deserves 12 pages.
\end{them}

Syntactic sugar plays an important role in achieving the desired level of
readability and understanding, so it is very important for achieving objective
(1a), which in turns supports the primary hypothesis. This is explained in the
beginning of \R{Section 12.3} , as well as in the \R{Chapter 16}, Results and
Evaluation (page ).

Operations are introduced at the beginning of \R{Chapter 11}, Main AF-Raf
Concepts, as requested by the examiner.

\begin{them}
Page 90:
``Unknown'' is named ``undefined'' in page 92, and only in page 93 there is the
statement that unknown and undefined are the same thing. Please choose one and
use it only.
\end{them}
Fixed. The symbol $\bot$ stands for 'undefined' or 'unknown', and both
references are consistent now. 

\begin{them}
Page 92:
The thesis states that ``agents states themselves are triples'', but in
definition 4.31 they are 6-tuples.  Such mismatches raise a lot of confusion
and a careful consistency check should be performed.
\end{them}
Fixed.

\begin{them}
Page 93:
``The sets AgentCode and GroundTerm are not defined here, but the definitions
can be easily recovered from the abstract grammar'': it should be better to
define them here, instead.
\end{them}
Fixed. The definition for agent code is presented in \R{figure 12.7}, and a
reference to the definition of ground term is provided.

\begin{them}
Page 95:
``P is the agent code'': the syntax of the agent code should be necessarily given
since this component is not clearly defined; in the reviewers' opinion, P is
not the actual agent code (meant as the program for an agent) but it is the
``agent's execution state'', which is a deeply different concept and requires
suitable data structures. The agent's code does not (usually) change, whereas
its excecution state does. Referring to the BDI model, the the AF-Raf agent
code can be compared with the plan library + the belief base, whereas the
execution state can be compared with the intentions set. Given the use of P
made in page 95 and the following ones, P seems closer to the BDI intentions
set. A clarification on the syntax and data structures used for P is required,
as well as a coherence check in all the transition rules.
\end{them}
The sentence ``$P$ is the agent code'' is correct.
I have added a formal definition of $P$ in \R{Figure~13.7}.
(I would note that the operational semantics of $\pi$-calculus is also
  often given in terms of `code'.
See, for example, the newly added rules \R{(9.5--12)}.
The message queues used there to represent asynchrony are `code'
  in the sense that they are used in the same places as code in the operational semantics.
In our case, things like a session identifier from \R{Figure~13.7}
  have an entirely snalogous role:
  programmers are not supposed to write such `code',
  but it is used in the operational semantics in exactly the same
    places as the code that \emph{can} be written by programmers.)


\begin{them}
Page 96:
Rule 4.44 states that we can discard I, O and d in the transition. It should be
stated where these components come into play again.
\end{them}
Fixed. Explanation added on \R{page 132}.

\begin{them}
Page 97:
The premise of rule 4.46 states that an agent's program is a set (since the
``belongs to P'' operator is used), which is not consistent with the definition
of ``agent code'' (for example, the agent code contains ``include'' statements and
it should include the beliefs as well). The consequence of the rule states that
an agent program is some data structure providing a ``;'' operator. This is in
contrast with stating that an agent's program is a set, as sets – as usually
defined - do not support a ``;'' operator, whatever the meaning of ``;'' is.
\end{them}
The new definition for agent code \R{(figure 12.7)} should clarify this.

Also, see \R{rule 12.52} for the definition of sequential composition.


\begin{them}
The rule also states that the agent's code changes over time, which supports
the reviewers' understanding that P is not the agent code, but the agent
execution state.
\end{them}
The agent execution state includes the agent code, as defined in figure 12.8. The agent execution state changes over time,
  including the agent code it contains.
(Of course, this does not change the agent code written by the programmer.)

We note that including agent code in the agent state is not unusual:
  Indeed, the usual way to give operational semantics to, say,
    $\pi$-calculus is as a transition system over fragments of `code'.

\begin{them}
Finally, the premise of the rule introduces a degree of nondeterminism. How are
rules actually selected for execution by the interpreter? A selection function
should be introduced, as in AgentSpeak(L). The candidate can stick to one
specific selection function for her implementation, but in the transition rules
the choice of the program rule to use should be demanded to a selection
function, for generality,
\end{them}
I disagree.
An explicit selection function is unnecessary for generality, and is ugly.
People have been using implicit nondeterminism,
  even in the sequential setting,
  since the dawn of operational semantics for nondeterministic languages.
For example,
  Plotkin handles nondeterminism this way in his semantics for
  Dijkstra's guarded commands;
see Plotkin, {\it Dijkstra's Predicate Transformers and Smyth' Powerdomains},
  1979.


\begin{them}
Page 99:
Rule 4.52: what does ``X:=P'' mean? Whatever P, assigning an agent code (which is
a set?) to a variable is not clear.
\end{them}
Rule 4.52 \R{(now 12.53)} is the definition of what $x:=P$ means, that is
assignment. Variable x is assigned the value to which the agent code is
evaluated to.

\begin{them}
Rule 4.53:
An agent code can be a belief? As Gamma t is a belief.
\end{them}
Yes. See agent code definition in \R{figure 12.7}.
The programmer is not allowed to write such code, though.
This is entirely analogous to the message queues in
  multiparty asynchronous session types:
  See the new background section covering them.

\begin{them}
``Here the scope Gamma is used as a substitution'': which are the relationship
with the notion of m.g.u.? Why not using the standard notation for
substitutions and m.g.u.s?
\end{them}
There is no unification here. I've changed the notation to the more usual
postfix.

\begin{them}
Page 100:
In the premise of rule 4.55 one element is missing as it should be a 6-tuple
\end{them}
Fixed.

\begin{them}
``The result of this substitution is the code that the newly created agent
should execute, and is a set of rules'': in Rule 4.53 it seemed that the agent
code could include beliefs as well.
\end{them}
Fixed. A ``set'' in this context has the meaning of a ``collection'' rather than
the mathematical concept of set. I've changed the formulation to avoid
confusion.

\begin{them}
Also, the comments to page 97 raise doubts that P can be a set because of the
``;'' operator.
\end{them}
See new agent code definition \R{(figure 12.7)} for clarification.

\begin{them}
Page 101:
``AF-Raf actions .. are an escape hatch from AF-Raf's semantics'': AF-Raf actions
should have a clean semantics as well; if it is not the case, a strong
motivation should be provided.
\end{them}
Programming languages can call pieces of code written in other programming
languages. For example, you can run Assembly code from C, and C code from Java
and Haskell. None of those programming languages have semantics for this
features.

\begin{them}
``If something is not possible to do in pure AF-Raf, ..., than an AF-Raf should
be used'': what does this sentence mean?
\end{them}
Fixed. ``(...) an AF-Raf action should be used.''

\begin{them}
Page 102:
The action send.... : what about the action receive? Is it supported? Which is
its semantics?
\end{them}
No, there's no receive action. Sensors deal with receiving messages in AF-Raf.

\begin{them}
Page 104:
Where was DetSessionState defined?
\end{them}
In rule 11.41

\begin{them}
In Rule 4.63, sid binds to nothing
\end{them}
The rule says that when sid doesn't match to any session it errors.

\begin{them}
Page 106:
Oneof, maybe, repeat have been introduced and explained nowhere.
\end{them}
Fixed. They are defined in syntax now.

\begin{them}
-- Chapter 5
The chapter shows the implementation of the same MAS in both Jason and AF-Raf,
but no systematic comparison is provided.
\end{them}
I have designed a questionnaire based survey to evaluate and compare the
readability and understanding of each implementation. I've considered the
individual programs, as well as how easy it is to put together related
programs. The results are presented in \R{Chapter 16}.

\begin{them}
-- Chapter 6
Page 118:
The sentence ``AF-Raf implementations tend to be shorter and more readable'' is
subjective. It may be true, but nothing in the thesis supports this statement.
\end{them}
Fixed. It is now supported by the survey results

\begin{them}
The sentence ``The AF-Raf programming language is properly formalised .... and
its operational semantics is defined...'' is not completely true in the current
version of the thesis: this is rather an objective that should be achieved in
the revised one.
\end{them}
Fixed. All is checked for correctness.

\begin{them}
Page 119:
The sentence ``...improved the efficiency...'' is subjective. Maybe it is true,
but nothing in the thesis supports this statement.
\end{them}
Efficiency is at least partly improved through type checking.
%}}} corrections
\end{document}
